---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

##         Information      

Author         :  <span style ="color :green"> Manas Chatterjee  </span> <br>
Date           :  March, 2019 <br>
Purpose        :  Week 2 project on Reproducible Research in Johns Hopkins University Data Science Courses 


## Loading and preprocessing the data

Install packages if not available   

```{r load_packages, echo=TRUE, warning=FALSE, message=FALSE}
# This block of code checks if the package is installed, and if not, installs it
list.of.packages <- c("dplyr", "tidyr", "RColorBrewer","ggthemes","ggplot2")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
```

Load packages in the environment  

```{r load_libraries, echo=TRUE, warning=FALSE, message=FALSE}
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ggthemes)
library(ggplot2)
```

Check if the file exists. If not download the file from given link 

```{r data_download,echo=TRUE,warning=FALSE, message=FALSE}
temp <- tempfile()
if (!file.exists('./data/files/activity.csv')) {
        print(paste0("You did not have the file; downloaded.... "))        
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
        unzip(temp,exdir = './data/files')
} 
```

Read the data into a dataframe `activity` 

```{r data_load,echo=TRUE,warning=FALSE, message=FALSE}
activity <- read.csv('./data/files/activity.csv', sep = ",",na.strings = "NA" )
# Now, use `str` to look at the structure of the dataframe
str(activity)
```

## What is mean total number of steps taken per day? 

```{r total_steps_per_day, echo = TRUE}
per_day <- activity%>%
    na.omit()%>%                            # removing the na data
    group_by(date)%>%
    summarise(total.steps=sum(steps),
              average.steps = mean(steps))
head(per_day)                               # This dataset contains MEAN total number of steps taken each day
```
Sneak peak into the dataset
 
```{r historgram_plot, echo=TRUE, message=FALSE,warning=FALSE}
g <- ggplot(per_day,aes(total.steps))
g + geom_histogram(binwidth = 1000, aes(fill=..count..))+
        stat_function(fun = dnorm, colour = "red")+
        scale_y_continuous(breaks=c(0,4,8,12))+
        scale_colour_economist()+
        labs(x="Steps taken in a single day",y="Count",title="Histogram of number of steps taken in a day ")
```
<br>Overall MEAN and MEDIAN

```{r mean_n_median, echo=TRUE,warnings=FALSE,message=FALSE}
# Overall mean - MEAN of average number of steps taken in a day
print(paste0(as.integer(mean(per_day$total.steps))))
# Overall median - MEDIAN of average number of steps taken in a day
print(paste0(median(per_day$total.steps)))
```

## What is the average daily activity pattern?

Finding MEAN steps for each interval

```{r interval_data, echo=TRUE,warnings=FALSE,message=FALSE}
interval_activity <- activity%>%
    na.omit()%>%                            # removing the na data
    group_by(interval)%>%
    summarise(average.steps = mean(steps))
```

Time series plot by interval

```{r max_interval, echo=TRUE,warnings=FALSE,message=FALSE}
l <- ggplot(interval_activity,aes(x=interval,y=average.steps))
l + geom_line(aes(colour=average.steps))+
#scale_x_continuous(breaks=seq(0,2500,by=250))+
#theme_economist()+
geom_text(aes(label=ifelse(interval_activity$average.steps == max(interval_activity$average.steps),
                           as.numeric(interval_activity$interval),'')))+
        labs(x="5 minute intervals",y="Average Steps",
             title="The busiest 5-minute interval")
```
<br>Result from the plot

```{r max_print, echo=TRUE}
print(paste0("The 5-minute interval with the maximum average steps is ",
interval_activity$interval[interval_activity$average.steps == max(interval_activity$average.steps)],
' with number of average steps ',as.integer(max(interval_activity$average.steps))))
```
## Imputing missing values

Checking for NA or missing values

```{r NA_columns, echo=TRUE}
sapply(activity,function(x) sum(is.na(x)))
```
Filling in missing values with MEAN by the same interval

```{r NA_replace, echo=TRUE}
new_activity <- activity%>%
    group_by(interval)%>%
    mutate(steps = ifelse(is.na(steps),mean(steps,na.rm=TRUE),steps))
    
filled_activity <- new_activity%>%
    group_by(date)%>%
    summarise(total.steps=sum(steps),
              average.steps = mean(steps))
    
    
head(filled_activity)                       # Summarized steps by day after filling in missing values
```

Histogram with new filled data (imputed).

```{r filled_hist, echo=TRUE}
g <- ggplot(filled_activity,aes(total.steps))
g + geom_histogram(binwidth = 1000, aes(fill=..count..))+
        stat_function(fun = dnorm, colour = "red")+
        labs(x="Steps taken in a single day",y="Count",
             title="Histogram of number of steps taken in a day (after filling in missing data)")
```

<br>Overall MEAN and MEDIAN

```{r mean_n_median_after, echo=TRUE,warnings=FALSE,message=FALSE}
# Overall mean - MEAN of average number of steps taken in a day
print(paste0(as.integer(mean(filled_activity$total.steps))))
# Overall median - MEDIAN of average number of steps taken in a day
print(paste0(as.integer(median(filled_activity$total.steps))))
```

## Are there differences in activity patterns between weekdays and weekends?

Convert date column into Date type

```{r factors, echo=TRUE}
new_activity$date <- as.Date(new_activity$date)
weekdays1 <- c('Monday',"Tuesday","Wednesday","Thursday","Friday")
new_activity$Weekday <- c('weekend','weekday')[(weekdays(new_activity$date) %in% weekdays1)+ 1L]
head(new_activity)
```

Checking the factors

```{r factor_check, echo=TRUE}
weekdays(new_activity$date[6])
filled_interval_activity <- new_activity%>%
    group_by(interval,Weekday)%>%
    summarise(average.steps = mean(steps))
head(filled_interval_activity)
```
Panel plot of activity comparing weekdays and weekends

```{r final_plot, echo=TRUE}
m <- ggplot(filled_interval_activity,aes(x=interval,y=average.steps, fill=Weekday))
m + geom_line(aes(colour=Weekday)) +
    facet_grid(Weekday~.)+ 
    labs(x="5 minute intervals",y="Average Steps",title="The busiest 5-minute interval (Weekend vs. Weekdays)")
```